\documentclass[12pt]{article}

\usepackage{theorem,amsmath,amssymb}

\def\Mor{\operatorname{Mor}}

\input{listki.tex}

\begin{document}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\listok{11}{GEOMETRY 11: Galois coverings}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

The subject of Galois covering that is covered in this exercise sheet
is very similar to the Galois theory of field extensions. This is not
a coincidence. In algebraic geometry methods from topology and
differential geometry are applied to objects of algebro-geometric and
number-theoretic nature. The version of Galois theory that is
presented in ALGEBRA-11 goes back to A. Grothendieck. Grothendieck has
given a definition of a fundamental group in such a way that Galois
group and fundamental group of a topological space turned out to be
particular cases of a more general construction. If one studies
coverings and field extensions, it is very useful to keep in mind that
these two things are similar.

All topological spaces in this exercise sheet are supposed to be
Hausdorff. 

%\begin{opredelenie}
%Пусть $M$ топологическое пространство, а
%$\tilde M \stackrel \pi \arrow M$ -- непрерывное отображение.
%Говорят, что $\pi$ {\bf этально}, если
%у каждой точки $\tilde M$ есть такая окрестность $U$,
%что ограничение $\pi$ на $U$ 
%осуществляет гомеоморфизм $U$ на $\pi(U)$.
%Иначе говоря, этальное отображение -- отображение,
%которое локально является гомеоморфизмом.
%\end{opredelenie}
%
%\begin{zadacha}
%Докажите, что этальное отображение открыто.
%\end{zadacha}

\begin{zadacha}
Let $\tilde M \stackrel \pi \arrow M$ be a covering
and let $M_1$ be a connected component of $\tilde M$.
Prove that $\pi^{-1}(M_1)$ is a connected component of  $\tilde M$. 
\end{zadacha}

\begin{zadacha}[!]
Let $\tilde M \stackrel \pi \arrow M$ be a covering and let $\tilde M$
and  $M$ be connected and non-empty,
and $\pi$ injective. Prove that $\pi$ is a homeomorphism.
\end{zadacha}

\begin{opredelenie}
  Let $\tilde M \stackrel \pi \arrow M$,
  $\tilde M'\stackrel {\pi'} \arrow M$ be coverings.  A {\bf morphism
    of coverings} is a continuous map
  $\phi:\; \tilde M\arrow \tilde M'$, that respects the projection to
  $M$, in other words, such that $\phi\circ \pi' = \pi$.  The set of
  all morphisms between coverings is denoted by
  $\Mor(\tilde M, \tilde M')$.  An isomorphism of covering is a
  morphism that is invertible, and moreover $\phi^{-1}\circ \phi=\Id$,
  $\phi\circ \phi^{-1}=\Id$.
\end{opredelenie}

\begin{zadacha}[!]
  Let $\phi:\; \tilde M\arrow \tilde M'$ be a morphism of
  coverings. Prove that $\phi:\; \tilde M\arrow \tilde M'$ is a
  covering.
\end{zadacha}

\begin{zadacha} 
Let $M$ be connected and $\tilde M \stackrel \pi \arrow M$ be a
covering. Prove that $\tilde M$ is locally connected.
\end{zadacha}

\begin{zadacha}
Let $M_1\arrow M_2$ and $M_2\arrow M_3$  be coverings.

\item[**] Is it true that the composition $M_1\arrow M_3$ is also a covering?

\item[!] Assume every point of $М_3$ has a simply connected neighbourhood.
Prove that  $M_1\arrow M_3$ is a covering.
\end{zadacha}

\begin{zadacha}
Let $\tilde M \stackrel \pi \arrow M$,
$\tilde M' \stackrel {\pi'} \arrow M$ be coverings and  $\tilde
M'\coprod \tilde M$ be their disjoint sum. Prove it is also a covering
of $M$.
\end{zadacha}

\begin{zadacha}
  Let $M$ be connected and $\tilde M \stackrel \pi \arrow M$ be a
  covering. Prove that 
  $\tilde M\cong \coprod_{\alpha\in I} \tilde M_\alpha$ where 
  $\{\tilde M_\alpha\}$ is the set of connected components of $\tilde
  M$ regarded as coverings of $M$.
\end{zadacha}

\begin{opredelenie}
  A {\bf splitting} of a covering $\tilde M \stackrel \pi \arrow M$ is
  an isomorphism between $\tilde M$ and a covering of the form
  $\tilde M \cong V \times M$ where $V$ is a set with discrete
  topologogy.
\end{opredelenie}

\begin{zadacha}
Let  $\tilde M \stackrel \pi \arrow M$ be a covering of a connected
space $M$. Prove that 
$\pi$ splits if and only if all connected components $\tilde M$ are
isomorphic to $M$.
\end{zadacha}


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subs{Galois coverings}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{zadacha}[!]
Let $M_1 \stackrel {\pi_1} \arrow M$,
$M_2 \stackrel {\pi_2} \arrow M$ be coverings.
Consider the following subset in $M_1\times M_2$
\[ M_1\times_M M_2:= 
\{ (m_1, m_2)\in M_1\times M_2 \ \ | \ \ \pi_1(m_1)= \pi_2(m_2)\}
\]
We consider $M_1\times_M M_2$ as a topological space (with the
topology induced from $M_1\times M_2$).  Prove that the natural map
$M_1\times_M M_2\arrow M$ is a covering.
\end{zadacha}

\begin{opredelenie}
  The space $M_1\times_M M_2$ together with the natural map to $M$ is
  called the {\bf product of coverings $M_1$, $M_2$}. The product of
  arbitrary number of coverings is defined similarly.
\end{opredelenie}

\begin{zamechanie}
  If one uses the analogy between field extensions and coverings then
  disjoint unions of coverings correspond to a direct sums of
  semisimple Artinian rings, and products of coverings correspond to
  tensor products.
\end{zamechanie}

\begin{zadacha} 
Let $M_1$, $M_2$, $M_3$ be coverings of $M$. Proe that
morphisms from $M_3$ to $M_1\times M_2$
are in bijective correspondence
with pairs of morphisms  $\phi_1:\; M_3\arrow M_1$, 
$\phi_2:\; M_2\arrow M_1$.
\end{zadacha}

\begin{zadacha}
Consider $\R$ as a covering of $S^1$.
How many connected components does $\R\times_{S^1}\R$ have?
\end{zadacha}

\begin{opredelenie}
Let $M_1\stackrel\phi\arrow M_2$ be a morphism between two coverings
of $М$. Define the  {\bf graph of $\phi$} as a subset in $M_1\times_M
M_2$ that consists of pairs of the form вида $(m, \phi(m))$ for all
$m\in M_1$.
\end{opredelenie}

\begin{zadacha}[!] 
Let $M_1\stackrel\phi\arrow M_2$ be a morphism between two coverings
$М$ and let $\Gamma_\phi$ be its graph. Prove that $\Gamma_\phi$ is
both open and closed in $M_1\stackrel\phi\arrow M_2$.
\end{zadacha}

\begin{zadacha}
  Let $[\tilde M:M]$ be a covering, and moreover let $M$ and
  $\tilde M$ be connected (such a covering is called {\bf connected}).
  Let $X\subset \tilde M\times_M\tilde M$ be a connected
  component. Prove that $X$ is the graph of an automorphism
  $\nu:\; \tilde M \arrow \tilde M$ if and only if the projection on
  the first components is an isomorphism $X\cong \tilde M$.
\end{zadacha}

\begin{zadacha}[!]\label{_Mor_v_pro_Zadacha_}
Let $[\tilde M:M]$ be a connected covering.
Consider the projection on the first argument
$\tilde M\times_M\tilde M \arrow \tilde M$ as a covering of 
$\tilde M$. Construct a bijective correspondence between
  $\Mor_{\tilde M}(\tilde M, \tilde M\times_M)$ and the set of
  automorphisms of $\tilde M$ over $M$.
\end{zadacha}

\begin{ukazanie}
Use the previous problem.
\end{ukazanie}

\begin{opredelenie} 
  Let $[\tilde M:M]$ be a covering and assume $M$ and $\tilde M$ are
  connected. Then $[\tilde M:M]$ is called a {\bf Galois covering}, if
  the covering $\tilde M\times_M\tilde M \arrow \tilde M$ is split. In
  this situation the automorphism group of $\tilde M$ over $M$ is
  called the {\bf Galois group of the covering $[\tilde M:M]$}
  (denoeted $\Gal([\tilde M:M])$). Sometimes the Galois group is
  called {\bf monodromy group}, or {\bf deck transform group}.
\end{opredelenie}

\begin{zadacha}[!]
Let $M$ be connected and let $[\tilde M:M]$ be such a Galois covering
that every point of $M$ has exactly $n$ preimages
(such a covering is called $n$-sheet covering). Prove that the Galois
group  $[\tilde M:M]$ has exactly $n$ elements.
\end{zadacha}

\begin{ukazanie} Prove thta $[\tilde M\times_M\tilde M:\tilde M]$ is
  $n$-sheet covering too, and use the previous problem.
\end{ukazanie}

\begin{opredelenie} Let a group $G$ act on a set $S$. The action is
  called {\bf free} if $s\neq gs$ for any $g\in G$, $s\in S$, if
  $g\neq 1$. The action is called {\bf transitive} if for any two
  points $s_1$, $s_2\in S$ there exists $g\in G$ such that
  $g(s_1)=s_2$.
\end{opredelenie}

\begin{zadacha}
Let $\tilde M \stackrel\pi\arrow M$ be a covering
and $G=\Aut_M(\tilde M)$ be its automorphism group. Assume that $M$ is
connected. Prove that for any  $x\in M$ the group $G$ acts freely on
$\pi^{-1}(x)$. 
\end{zadacha}

\begin{zadacha}[!]
Let $\tilde M \stackrel\pi\arrow M$ be a Galois covering and let $x\in
M$ be any point. Prove that $\Gal([\tilde M:M])$ acts on $\pi^{-1}(x)$
freely and transitively.
\end{zadacha}

\begin{ukazanie}
Find a bijective correspondence between
 $\pi^{-1}(X)$ and the set of connected componentns
$\tilde M\times_M\tilde M$, and apply
Exercise~\ref{_Mor_v_pro_Zadacha_}. 
\end{ukazanie}

\begin{zadacha}[!]\label{_transi_na_sloe_Zadacha_}
Let $\tilde M \stackrel\pi\arrow M$ be a covering and let $x\in M$ be
any point. Prove that $\Aut_M(\tilde M)$ acts transitively on
$\pi^{-1}(X)$ if and only if  $[\tilde M: M]$ is a Galois covering. 
\end{zadacha}

\begin{zadacha} 
  Consider the covering $\R^n \arrow \R^n/\Z^n \cong (S^1)^n$.  Prove
  that it is a Galois covering.
\end{zadacha}

\begin{zadacha} 
  Take $n\in \Z$ and consider an $n$-sheeted covering
  $S^1 \arrow S^1$, $t \mapsto nt$.  Prove that this is a Galois
  covering.
\end{zadacha}

\begin{opredelenie} Let $M$ be a topological space and let 
$G$ be a group that acts on $M$ by continuous transformations. 
Consider the space of $G$-orbits
$M/G$. Recall (GEOMETRY-10) that the topology on $M/G$ is introduced a
follows: a subset of $M/G$ is open if and only if 
its preimage in  $M$ is open. The set $M/G$ with this topology is called
a  {\bf quotient of} $M$ by the action of $G$.
\end{opredelenie}

\begin{zadacha}[!]
  Let $[\tilde M:M]$ be a covering and assume
  $G\subset \Aut_M(\tilde M)$ acts on $[\tilde M:M]$ by
  automorphisms. Prove that this action is free and that the quotient
  $\tilde M/G$ is Hausdorff and is a covering of $M$.
\end{zadacha}

\begin{zamechanie}
Taking a quotient by the action of $G$ plays the same role in the
Galois coverings theory as taking  $G$-invariant in the Galois theory
of field extensions.
\end{zamechanie}

\begin{zadacha}[!]
  Let $[\tilde M:M]$ be a covering and let $G$ be its automorphism
  group. Prove that $\tilde M/G$ is isomorphic to $M$ if and only if
   $[\tilde M:M]$ is a Galois covering. 
\end{zadacha}

\begin{ukazanie}
Use Exercise~\ref{_transi_na_sloe_Zadacha_}.
\end{ukazanie}

\begin{zamechanie}
  In the several exercises that follow the statement and the proof
  mimic almost verbatim the corresponding exercises about Galois field
  extensions.
\end{zamechanie}

\begin{zadacha}\label{_kompo_rasshe_Zadacha_}
Let $M_1\stackrel{\phi_1}\arrow M_2 \stackrel{\phi_2}\arrow M_3$
be a sequence of coverings, and moreover
$\phi_i$ are surjective and their composition is split.
Prove that $\phi_i$'s split.
\end{zadacha}

By analogy with Galois theory of field extension
the coverings of the form $\tilde M \stackrel \pi \arrow M$ will
further be denoted $[\tilde M:M]$.


\begin{zadacha}[!]\label{_M_1_times_promezhu_Zadacha_}
  Let $M_1\arrow M_2 \arrow M_3$ be a sequence of coverings, and
  assume all $M_i$ are connected and $[M_1:M_3]$ is a Galos
  covering. Prove that $M_1 \times_{M_3} M_2$ splits as a covering of
  $M_1$.
\end{zadacha}

\begin{ukazanie} 
Use Exercise~\ref{_kompo_rasshe_Zadacha_}, apply it to 
the sequence of coverings
\[ 
  M_1\times_{M_3}M_1\arrow M_1\times_{M_3}M_2 \arrow M_1\times_{M_3}M_3.
\]
\end{ukazanie}

\begin{zadacha}[!]
  Let $M_1\arrow M_2 \arrow M_3$ be a sequence of coverings and assume
  that $[M_1:M_3]$ is a Galois covering. Prove that $[M_1:M_2]$ is a
  Galois covering.
\end{zadacha}

\begin{ukazanie}
Use Exercise~\ref{_kompo_rasshe_Zadacha_}.
\end{ukazanie}

\begin{zadacha} 
  Let $M_1\arrow M_2 \arrow M_3$ be a sequence of coverings. Prove
  that
\[ M_1\times_{M_3} M_1 \cong 
   M_1\times_{M_2} (M_2\times_{M_3}M_2)\times_{M_2}M_1.
\]
\end{zadacha}

\begin{zadacha} Deduce the following statement from the previous exercise:
if $M_1\arrow M_2 \arrow M_3$ is a sequence of coverings and if $[M_1:M_2]$
and $[M_2:M_3]$ is a Galois covering then $[M_1:M_3]$ is also a Galois
covering. 
\end{zadacha}

\begin{zadacha} 
Let $[\tilde M:M]$ be a covering and let 
$G$ be its Galois gorup and $G'\subset G$ be its subgroup. Consider
the quotient $\tilde M/G'$. Prove that
$[\tilde M: \tilde M/G']$ is a Galois covering with Galois group
$G'$. 
\end{zadacha}

\begin{opredelenie}
Let $\tilde M \arrow M$ be a covering.
A {\bf quotient covering} $[\tilde M:M]$ is a covering $\tilde
M'\arrow M$ together with a sequence of coverings
$\tilde M \arrow \tilde M' \arrow M$.
\end{opredelenie}

\begin{zadacha}[!]
(fundamental theorem of Galois theory)
Let $[\tilde M:M]$ be a Galois covering with Galois group
$G$. Consider the correspondence that to a subgroup $G'\subset G$
associates a quotient covering $[\tilde M/G':M]$.
Prove that this correspondence defines a bijection between the set of
subgroups of $G$ and the set of isomorphism classed of quotient
coverings. 
\end{zadacha}

\begin{zadacha} 
Let $M_1\arrow M_2 \arrow M_3$ be a sequence of coverings and assume
that $[M_1:M_3]$ is a Galois covering. Consider the natural projection 
\[ M_1 \times_{M_3} M_1 \stackrel\Psi\arrow  M_2 \times_{M_3} M_2.
\]
Let $g\in \Gal([M_1:M_3])$ and $e_g\subset M_1 \times_{M_3} M_1$ be
the graph 
$\{(m, g(m))\}$ of the action of $g$ in $M_1 \times_{M_3} M_1$.
Prove that  $g\in \Gal([M_1:M_2])\subset \Gal([M_1:M_3])$
if and only if $e_g$ projects to the diagonal componont in 
$M_2 \times_{M_3} M_2$.
\end{zadacha}

\begin{zadacha} 
Let $M_1\arrow M_2 \arrow M_3$ be a sequence of Galois
coverings. Prove that the natural projection
\[ M_1 \times_{M_3} M_1 \stackrel\Psi\arrow  M_2 \times_{M_3} M_2.
\]
defines a surjective homomorphism
$\Gal([M_1:M_3])\stackrel\psi\arrow \Gal([M_2:M_3])$. 
Prove that $\ker\psi=\Gal([M_1:M_2])$.
\end{zadacha}

\begin{ukazanie}
Use the fact the Galois $\Gal([M_i:M_3])$ is identified with the set
of connected components of 
$M_i\times_{M_3}M_i$ and use the previous problem.
\end{ukazanie}

\begin{zadacha}[!]
Let $\tilde M \arrow M$ be a Galois covering and  $G' \arrow \tilde
M/G'$ be the bijective correspondence between quotient coverings and
subgroups of the Galois gorup defined above.
Prove that  $G'$ is a normal subgroup if and only if $[\tilde M/G':M]$
is a Galois covering.
\end{zadacha}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subs{Coverings of linearly connected spaces}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{opredelenie}
Let $M$ be a metric space. Recall that a 
{\bf geodesic} in $M$ is a path
$[a,b] \stackrel\gamma\arrow M$ such that $d(\gamma(x), \gamma(y))=|x-y|$.
The {\bf length} of the geodesic is the distance between its ends. A
path is called {\bf piece-wise geodesic} if it can be decomposed into
a union of a finite number of geodesic segments. The {\bf length} of a
piece-wise geodesic path is defined to be  the sum of lengths of its
geodesic pieces. We denote the length of a path  $\gamma$ by $|\gamma|$.
\end{opredelenie}

\begin{opredelenie}
Let $\Gamma$ be a graph and $M_\Gamma$ be its topological space.
We say that $\Gamma$ is {\bf  connected} if its topological space is
connected.  
\end{opredelenie}

\begin{zadacha}[!]
Prove that a graph is connected if and only if any two vertices are
connected by a finite sequence of edges. Prove that a connected graph
is linearly connected.
\end{zadacha}

\begin{zadacha}[!]
  Let $\Gamma$ be a connected graph. By construction, on each edge
  $r_\alpha\subset M_\gamma$ of the graph are defined the coordinates
  that identify the edge with $[0,1]$.  Let $\gamma$ be a piece-wise
  linear path in $\Gamma_M$, that is, a path that consists of a finite
  number of intervals of the form
  $[a_i,b_{i}]\stackrel{\phi_i}\arrow [\lambda_i, \mu_i]\subset
  r_\alpha$, where $\phi_i$ is linear. Define 
  $|\gamma|:= \sum|\lambda_i,\mu_i|$ as the sum of lengths of all
  intervals that contain this path. Define $d(x,y):= \inf |\gamma|$ where
  $\gamma$ runs throuh all piece-wise linear paths from $x$ to $y$.
  Prove that $d(x,y)$ defines a metric and $M_\Gamma$ is geodesically
  connected. 
\end{zadacha}

\begin{opredelenie}
  This metric is called the {\bf standard metric on the topological
    space of a graph}.
\end{opredelenie}

\begin{opredelenie}
Geodesically connected manifold $M$ is called
{\bf star-shaped} if any two points of $M$ are connnected by a unique
geodesic.
\end{opredelenie}

\begin{zadacha} 
  Prove that any convex subset in $\R^n$ (with the standard metric) is
  star-shaped.
\end{zadacha}

\begin{zadacha}[*]
Find a metric on $M=\R^2$ such that  $М$ is geodesically connected and
there are infinitely many geodesics connecting arbitrary two fixed points.
\end{zadacha}

\begin{zadacha}[*]
Let $\Gamma$ be a tree, that is, a connected finite graph that has 
$n$ vertices and $n-1$ edges. Prove that 
$M_\Gamma$ with the standard metric is start-shaped.
\end{zadacha}

\begin{zadacha}[*]
Let $\Gamma$ be a finite graph such that $\Gamma_M$
is star-shaped. Prove that $\Gamma$ is a tree.
\end{zadacha}

\begin{zadacha}[!]
Let $M$ be a geodesically connected manifold, 
$\tilde M\stackrel\pi\arrow M$ be a covering, and $x$ and $y$ be two
points in $\tilde M$.  Consider the set $S_{x,y}$ of all paths on
$\tilde M$ that connect $x$ and $y$, such that their projection to $M$
is piece-wise geodesic.
Consider the following function on $\tilde M \times \tilde M$
\[ \tilde d(x,y) = \inf_{\gamma\in S_{x,y}}|\pi(\gamma)|\] Prove that
it is a metric. Prove that $\tilde d(x,y)\geq d(\pi(x),\pi(y))$.
\end{zadacha}

\begin{zadacha}[*]
In the previous problem setting prove that
$\tilde M$ is geodesically connected.
\end{zadacha}

\begin{zadacha} Let $M$ be a geodesicall connected metric space and
  $\tilde M\arrow M$ be its covering. Prove that the connected
  component of the preimage of a geodesic is a geodesic in
  $(\tilde M, \tilde d)$.
\end{zadacha}

\begin{ukazanie} 
Prove that the preimage of a geodesic is a geodesic in a neighbourhood
of every point. Then use the inequality 
$\tilde d(x,y)\geq d(\pi(x),\pi(y))$.
\end{ukazanie}

\begin{zadacha}[!]
  Let $(M, d)$ be a star-shaped metric space and
  $\tilde M \stackrel\pi \arrow M$ be its connected covering. Let
  moreover $x\in \tilde M$ be any point and $U_x$ be the set of points
  $y\in M$ that can be connected with $x$ by a geodesic. Prove that
  $U_x$ is open and closed in $\tilde M$ and that $(U_x, \tilde d)$ is
  star-shaped. Deduce that the natural projection 
  $\tilde M \stackrel\pi \arrow M$ is an isometry and a homeomorphism.
\end{zadacha}

\begin{ukazanie}
Use the previous exercise.
\end{ukazanie}

\begin{zadacha}
  Let $M=[0,1]\times [0,1]$ be a square and $\tilde M \arrow M$ be its
  connected covering. Prove that it is a homeomorphism.
\end{zadacha}

\begin{zadacha}
Let $M$ be a linearly connected and simply connected space, 
and $\tilde M \stackrel\pi\arrow M$ be a connected covering. Prove that
it is a homeomorphism
\end{zadacha}

\begin{ukazanie}
  Prove that $\tilde M$ is linearly connected.  Let
  $x, y\in \pi^{-1}(x_0)$ be two points and $\tilde \gamma$ be a path
  that connects them.  Then $\gamma:=\pi(\tilde\gamma)$ is a
  loop. Since $M$ is simply connected, $\gamma$ can be extended to a
  map from the square to $X\subset M$ (prove it).  Consider the
  preimage of this square in $\tilde M$ and let $\tilde X$ be the
  component of the preimage that contains $\tilde\gamma$.  Use the
  previous exercise to prove that $\tilde X\stackrel\pi\arrow X$ is a
  homeomorphism and deduces that $x=y$.
\end{ukazanie}

\begin{zadacha}
In the previous problem setting prove that any covering $M$ splits.
\end{zadacha}

\begin{opredelenie}
Let $M$ be a any (not necessarily linearly connected) connected
topological space. The space $M$  is called 
 {\bf simply connected} if any covering of $M$ is split.
\end{opredelenie}

\begin{zamechanie} 
Thanks to the previous exercise this definition is consistent with the
definition of a simply connected linearly connected topological spaces
given in GEOMETRY 10.
\end{zamechanie}

\begin{opredelenie}
Let $M$ be connected. A covering  $\tilde M \arrow M$  is called  {\bf
  universal} if it is simply connected.
\end{opredelenie}

\begin{zadacha}[!] Prove a universal covering is a Galois covering. 
\end{zadacha}

\begin{zadacha}[!]
Prove that universal covering is unique up to isomorphism.
\end{zadacha}

\begin{ukazanie} Let $\tilde M$, $\tilde M'$ be two universally
  coverings of $M$. Since $\tilde M\times_M\tilde M'$ is a covering of
  $\tilde M$, $\tilde M'$, it splits over $\tilde M$, $\tilde
  M'$. This means that any connected component
  $\tilde M\times_M\tilde M'$ projects isomorphically to $\tilde M$,
  $\tilde M'$.
\end{ukazanie}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subs{Existence of the universal covering}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{zadacha}
Let $M$ be linearly connected,
$\tilde M \overset{\pi}{\arrow} M$ be a connected covering and 
$x\in M$ be any point.
Prove that the cardinality of the set $\pi^{-1}(x)$  is not greater
than the cardinality of $\pi_1(M)$.
\end{zadacha}

\begin{zadacha}
Prove that
the cardinality  $\pi^{-1}(x)$ is not greater than the cardinality of
the set $M^{[0,1]}$ of maps from  $[0,1]$ to $M$.
\end{zadacha}

\begin{zadacha}[*]
  Let $\tilde M \overset{\pi}{\arrow} M$ be a connected covering of a
  connected $M$ and $x\in M$ be any point. Prove that the cardinality
  of $\pi^{-1}(x)$ is not greater than $|2^{2S}|$, where $|2^{2S}|$ is
  the cardinality of the set of subsets of $S\times S$.
\end{zadacha}

\begin{ukazanie}
Choose $x_1, x_2\in \pi^{-1}(x)$. Prove that there exists a collection
of such connected open subsets $\{\tilde U_\alpha\}\in \pi^{-1}(S)$
that  $\tilde U_{\alpha_0}$ has non-empty intersection with the union
of all  $\tilde U_\alpha$ that are not equal to $U_{\alpha_0}$, and
moreover 
\[ \{x_1 ,x_2\}= \pi^{-1}(x)\cap (\bigcup \tilde U_\alpha)\]
Decreasing the base $S$ if necessary one can assume that
 $\pi$ splits over  $\pi(U_\alpha)$ for all
$\alpha$. Prove that $x_2$ is determined uniquely if  $x_1$,
$\{\pi(U_\alpha)\}$ is given, and if it is known which $U_{\alpha}$
have non-empty intersection.
\end{ukazanie}

\begin{zadacha} 
Let $M$ be connected and let $V$ be a set of a cardinality defined below.
Denote by ${\cal R}$ the set of all topologies defined on some subset  
$X\subset M \times V$ in such a way that 
the only natural projection $X\arrow M$
is a covering. Prove that any connected covering $M$ is isomorphic to
some element of  ${\cal R}$ if 
\begin{enumerate}
\item $M$ is linearly connected and the cardinality of $V$ 
is $|M^{[0,1]}|$
\item\sttr{} The cardinality of
$V$ is $|2^{2S}|$ where $S$ is the base of topology on  $M$.
\end{enumerate}
\end{zadacha}

\begin{zamechanie}
This exercise allows one to speak of  ``the set of isomorphism classes
of coverings''. Recall that not all mathematical objects are sets;
thus, the class of all sets is not a set. In order to prove that a
class is a set, one has to restrict its cardinality.
\end{zamechanie}

\begin{opredelenie}
Let $\{M_\alpha\stackrel{\pi_\alpha}\arrow M\}$ be a collection of
maps onto $M$ indexed b $I$ (possibly infinite, and even
uncountable). Consider the set of all  
$(m_{\alpha_1}, m_{\alpha_1},\dots)\in \prod M_{\alpha}$ such that
$\pi_{\alpha}(m_\alpha)=m$ for some
$m\in M$. This set is called the {\bf fibre product of
  $\{M_{\alpha}\}$} and is denoted by
$\prod_M M_{\alpha}$.
\end{opredelenie}

\begin{zadacha}
Let $M$ be a topological space and
$\{M_\alpha\stackrel{\pi_\alpha}\arrow M\}$ be a collection of its
coverings. Introduce on  $\prod_M M_{\alpha}$ a topology in the
following way. Let
$U\subset M$ be open and let $\{U_\alpha\subset M_{\alpha}\}$ be the
collection of open sets that cover $U$. Prove that the sets of the
form $\prod_U U_\alpha\subset \prod_M M_{\alpha}$ define the base of
topology on $\prod_M M_{\alpha}$. Prove thta $\prod_M M_{\alpha}$ is
Hausdorff.

\item[*] Is it true that the natural projection
$\prod_M M_{\alpha}\arrow M$ is a covering?

\item[!] Suppose that every point of $М$ has a simply connected
  neighbourhood. Prove that the natural projection $\prod_M
  M_{\alpha}\arrow M$ is a covering.

\end{zadacha}

\begin{opredelenie}
In this situation $\prod_M M_{\alpha}$ is called
the {\bf fibred product of $M_{\alpha}$ over $M$} or just a product of
coverings 
$M_\alpha\stackrel{\pi_\alpha}\arrow M$.
\end{opredelenie}

\begin{zadacha} \label{_rascheplya_product_Zadacha_}
Assume all coverings $\{M_\alpha\stackrel{\pi_\alpha}\arrow M\}$ are
split. Prove that  $\prod_M M_{\alpha}$ split too.
\end{zadacha}

\begin{zadacha}[!]
Let $\{M_\alpha\stackrel{\pi_\alpha}\arrow M\}$ be a Galois
covering. Prove thta any connected component of their product over $M$
is a Galois covering too.
\end{zadacha}

\begin{ukazanie}
Use the Exercise~\ref{_rascheplya_product_Zadacha_}.
\end{ukazanie}

\begin{zadacha} 
Let $\tilde M$ be a covering of $M$.
Constructi the natural bijection between 
$\Mor(\prod_M M_{\alpha}, \tilde M)$
and $\prod \Mor(M_{\alpha}, \tilde M)$
\end{zadacha}

\begin{zadacha}[*]
Let $\{M_\alpha\stackrel{\pi_\alpha}\arrow M\}$ be the set of all
coverings $S^1 \arrow S^1$, $t\arrow nt$,
indexed by $n\in \Z$. Prove that any
connected component of $\prod_M M_{\alpha}$ is isomorphic to 
$\R\arrow \R/\Z = S^1$.
\end{zadacha}

\begin{zadacha} 
  Let $\tilde M\stackrel\pi\arrow M$ be a covering, and assume that
  $\tilde M$ and $M$ are connected, $x\in M$,
  $x_1, x_2 \in \pi^{-1}(x)$, $W$ is the connected component of
  $\tilde M\times_M\tilde M$ that contains $x_1\times x_2$, and $W_1$
  is the connected component of
  $\tilde M\times_M\tilde M\times_M\tilde M$, that contains
  $x_1\times x_2\times x_2$. Prove that the natural projection
  $W_1 \arrow W$ (forgetting the third argument) is an isomorphism.
\end{zadacha}

\begin{zadacha} 
  In the same situation, let $\{x_\alpha\}$ be a set of points in
  $\pi^{-1}(x)$, indexed by $\alpha\in I$, and let $W$ be the
  corresponding component in the fibre product $\prod_{M,I}\tilde M$
  of $I$ copies of $\tilde M$, and $W_1$ be a component in
  $\left(\prod_{M,I}\tilde M\right)\times_M\tilde M$, that contains
  $\{x_\alpha\}$ and $x_0$, and moreover $x_0\in \{x_\alpha\}$.  Prove
  that the natural projection $W_1 \arrow W$ is an isomorphism.
\end{zadacha}

\begin{zadacha}[!]
  Let $\tilde M \stackrel \pi \arrow M$ be a connected covering and
  $x\in M$. Consider the product $\prod_{M,\{\pi^{-1}(x)\}}\tilde M$
  of $\tilde M$ with itself indexed by the set $\pi^{-1}(x)$, and let
  $\tilde M_G$ be the connected component in
  $\prod_{M,\{\pi^{-1}(x)\}}\tilde M$ containing the product of all
  $x_\alpha \in \{\pi^{-1}(x)\}$.  Prove that 
  $\tilde M_G\times_M \tilde M$ splits over $\tilde
  M_G$. Prove that $\tilde M_G \arrow M$ is a Galois covering.
\end{zadacha}

\begin{zamechanie}
We have proved that any covering is a quotient covering of a Galois
covering. 
\end{zamechanie}

\begin{zadacha}  
Let $M$ be a connected topological space,  
${\cal R}$ be the set of all isomorphism classes of connected
coverings of  $M$, and let $\{M_\alpha\stackrel{\pi_\alpha}\arrow M\}$
be the corresponding set of coverings, and 
$\tilde M\subset \prod_M M_{\alpha}$ be the connected component of
their product. Prove that for any connected covering  $\tilde M'
\arrow \tilde M$ there exists a surjective morphism of coverings
$\tilde M \arrow \tilde M'$. 
\end{zadacha}

\begin{ukazanie} Use the previous exercise.
\end{ukazanie}

\begin{zadacha}
In the previous problem setting prove that $\tilde M$ is a Galois
covering. 
\end{zadacha}

\begin{zadacha}[!]
Deduce tht for any $\tilde M \arrow M$ the covering 
$\tilde M \times_M \tilde M'\arrow \tilde M$ splits.
\end{zadacha}

\begin{ukazanie} Use the Exercise~\ref{_M_1_times_promezhu_Zadacha_}.
\end{ukazanie}

\begin{zadacha}[!]
Let $M$ any connected topological space,
$\tilde M \arrow M$ be a Galois covering constructed above.
Prove that $\tilde M$ is simply connected.
\end{zadacha}

\begin{zamechanie}
We have obtained that any connected topological space has a universal
covering.  
As was shown above, the universal cover is unique.
\end{zamechanie}

\begin{zadacha}[!]
Let $M$ be linearly connected, and $\tilde M$ be its universal
covering, and $\Gal([\tilde M:M])$ be the corresponding Galois
group. Prove that $\Gal([\tilde M:M])$ is not isomorphic to the Galoi
group of $M$.
\end{zadacha}

\begin{opredelenie}
The {\bf fundamental group} of a topological space $М$ is the group
$\pi_1(M):=\Gal([\tilde M:M])$, where  $\tilde M$ is the universal
covering. 
\end{opredelenie}

\begin{opredelenie} 
Subgroups $G_1, G_2\subset G$ are called {\bf
conjugated} if there exists $g\in G$
such that $G_1$ is mapped to $G_2$ by the automorphism 
$x\arrow x^g$.
\end{opredelenie}

\begin{zadacha}[*]
Let $M_1\arrow M$ be a covering, and let
$\tilde M \arrow M_1\arrow M$ be the universal covering.
Consider the subgroup  $G_1\subset \Gal([\tilde M:M])=\pi_1(M)$, 
obtained as a result of the fundamental theory of the Galois
theory. Prove that this correspondence defines a bijection
between isomorphism classes of coverings of $M$
and conjugacy classes of subgroups of  $\pi_1(M)$. 
\end{zadacha}

\begin{zadacha}[!]
Find all coverings of a circle up to isomorphism. Construct them explicitly.
\end{zadacha}

\begin{zadacha}[*]
Let $M$ be a connected topological space such that all linear
connected components of it are simply connected. Can it have a
non-trivial fundamental group?
\end{zadacha}

\begin{zadacha}[*]
Let $B$ be the set of polynomials
$P(t)= t^n+ a_{n-1}t^{n-1}+ a_{n-2}t^{n-2}+ \dots + a_0$
over $\C$ that have distinct roots and let $B_1$ be the set of all
tuplese $(x_1,\dots, x_n)\in \C^n$ of pairwise distinct numbers
$x_i \in \C$. Introduce on $B$ and $B_1$ the natural topology of a
subset of $\C^n$. Consider the map
$B_1\stackrel\pi\arrow B$, $(x_1,\dots, x_n)\arrow \prod(t-x_i)$.
Prove that  $\pi$ is a Galois covering. Find its Galois group.
\end{zadacha}

\begin{zadacha}[*]
Construct a connected covering that is not a Galois cover.
\end{zadacha}

\end{document}

%%% Local Variables: 
%%% mode: latex
%%% coding: koi8-r
%%% ispell-local-dictionary: "british"
%%% TeX-master: t
%%% End: 
